---
title: "DescrTab Tutorial"
output: rmarkdown::pdf_document
header-includes:
  - \usepackage{longtable}
  - \usepackage{geometry}
  - \usepackage{pdflscape}
  - \newcommand{\blandscape}{\begin{landscape}}
  - \newcommand{\elandscape}{\end{landscape}}
vignette: >
  %\VignetteIndexEntry{DescrTab Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

`DescrTab` is a package to create descriptive statistics tables.
It supports a variety of different customization options and can be used
in .Rmd files in conjunction with knitr.

## Preamble settings

### imbi_report
You're all set. Everything is already included.

### pdf_document
If you want to use `DescrTab` inside a regular .Rmd file, this is an example of a working preamble:

````markdown
---
title: "DescrTab Tutorial"
output: pdf_document
header-includes:
  - \usepackage{longtable}
---
````

### word_document
No special preamble needed. Make sure you have pandoc version >= 2.0 installed on your system.


## Basic examples for use in pdf_document

### Hello world

Make sure you include the DescrTab library by typing 

```{r, echo=TRUE, message=FALSE}

library(DescrTab)

```
somewhere in the document before you use it.

Creating a simple descriptive statistics table now is as easy as typing:

```{r, results='asis'}
des.print(madeUpData)
```
```

```{r, echo=FALSE, results='asis', message=FALSE, warning=FALSE}

des.print(madeUpData)

```

Note the chunk option `results='asis'`. 
The `des.print` function returns \LaTeX code by default. 
The directive `results='asis'` tells knitr to interpret the output of des.print 
as if it were pasted into the .tex file that is used to create the pdf output. 
Without the `results='asis'` chunk option, the table will not be rendered 
properly in the output document.

### The most common usecase

Usually a study population is divided into serveral groups of interest, and 
we would like to examine a comparison of these.
This can be achieved by using the `group` parameter:


```{r, echo=TRUE, results='asis', message=FALSE, warning=FALSE}

des.print(madeUpData, group="group")

```

You can assign labels to the different variables using the ` var.names ` parameter.
Additionally, setting ` data.names=F ` will surpress output of the variable names
and only print out the given labels.

```{r, echo=TRUE, results='asis', message=FALSE, warning=FALSE}

des.print(madeUpData, group="group",
    var.names=c("Sex", "Age", "Hq in Blood in mg/L", "Quality of Life"), data.names=F)

```


You can change the caption of the table using the ` tab.caption ` parameter.
The labels of the grouping variables can be controlled via the ` caption ` parameter.

```{r, echo=TRUE, results='asis', message=FALSE, warning=FALSE}

des.print(madeUpData, group="group", tab.caption="A nice table", caption=c("Placebo", "Experimental"))

```


### Help: My table is too big for the page (horizontally).

There are a couple of mechanisms that can be used to deal with this problem.

Firstly, you can try to decrease the fontsize via the ` fsize ` parameter 
(default is ` fsize = 11 `), which will shrink the whole table:

```{r, echo=TRUE, results='asis', message=FALSE, warning=FALSE}

des.print(madeUpData, group="group", fsize=6)

```
\newpage

\blandscape
\thispagestyle{empty}

You could also try temporarily switching your document to landscape mode. 
This can by done by including the following in your preamble:

````rmarkdown
header-includes:
  - \usepackage{longtable}
  - \usepackage{pdflscape}
  - \newcommand{\blandscape}{\begin{landscape}}
  - \newcommand{\elandscape}{\end{landscape}}
````

You can now enter landscape mode by typing:

````rmarkdown
\newpage
\blandscape
This is some text i want to have in landscape mode
\elandscape

````
Appropriate \LaTeX magic may be applied to turn the header and footer row, see e.g. \\ https://tex.stackexchange.com/questions/278113/single-landscape-page-with-page-number-at-the-bottom \\

An alternative way, for the imbi_report document type:
````rmarkdown
header-includes:
- \usepackage{typearea}
- \usepackage{fancyhdr}
- \fancypagestyle{lscape}{\headwidth\textwidth}


…..<nächste Seite soll umgedreht werden>
\clearpage 
\KOMAoptions{twoside=false}
\KOMAoption{paper}{landscape}
\typearea{12}
\pagestyle{lscape}

…..

…..<nächste Seite soll wieder normal sein >
\clearpage
\KOMAoption{paper}{portrait}
\typearea{12}
````
\vfill
\raisebox{0cm}{\makebox[\linewidth]{\thepage}}
\elandscape


