---
title: "Usage guidance"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Usage guidance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(knitr)
library(tidyverse)
library(magrittr)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

`DescrTab2` is the replacement of the `DescrTab` package.
It supports a variety of different customization options and can be used
in .Rmd files in conjunction with knitr.

## Preamble settings
`DescrTab2` works in your R-console, as well as in `.Rmd` documents corresponding to output formats
of the type `pdf_documument`, `html_document` and `word_document`.
For example, if your YAML-header looks like the example below, `DescrTab2` should automagically detect the output format and possibly load the required LaTeX packages.


````markdown
---
title: "DescrTab2 tutorial"
output: pdf_document
---
````

## Getting started
Make sure you include the DescrTab2 library by typing 

```{r, echo=TRUE}
library(DescrTab2)
```
somewhere in the document before you use it. You are now ready to go!


For instructive purposes, we will use the following dataset:
```{r}
dat <- iris[, c("Species", "Sepal.Length")]
dat %<>% mutate(animal= c("Mammal", "Fish") %>% rep(75) %>% factor())
dat %<>% mutate(food= c("fries", "wedges") %>% sample(150, TRUE) %>% factor())
```


Producing beautiful descriptive tablesis now as easy as typing:

```{r, results='asis', echo=F}
descr(dat)
```


## Accessing table elements

The object returned from the `descr` function is basically just a named list. You may be interested in referencing certain summary statistics from the table in your document. To do this, you can save the list returned by `descr`:

```{r}
my_table <- descr(dat)
```

You can then access the elements of the list using the `$` operator.

```{r}
my_table$variables$Sepal.Length$results$Total$mean
```

Rstudios autocomplete suggestions are very helpful when navigating this list.

The `print` function returns a formatted version of this list, which you can also save and access using the same syntax.

```{r, results='asis'}
my_table <- descr(dat) %>% print(silent=TRUE)
```


## Specifying a group
Use the `group` option to specify the name of a grouping variable in your data:

```{r, results='asis'}
descr(dat, "Species")
```

## Assigning labels
Use the `group_labels` option to assign group labels and the `var_labels` option to assign variable labels:

```{r, results='asis'}
descr(dat, "Species", group_labels=list(setosa="My custom group label"), var_labels = list(Sepal.Length = "My custom variable label"))
```

## Confidence intervals for two group comparisons
For 2-group comparisons, decrtab automatically calculates confidence intervals for differences in effect measures:

```{r, results='asis'}
descr(dat, "animal")
```

## Different tests
There are a lot of different tests available. Check out the test_choice vignette for details: https://imbi-heidelberg.github.io/DescrTab2/articles/test_choice_tree_pdf.pdf

Here are some different tests in action:

```{r, results='asis'}
descr(dat %>% select(-"Species"), "animal", test_options = list(exact=TRUE, nonparametric=TRUE))
```


```{r, results='asis'}
descr(dat %>% select(c("Species", "Sepal.Length")), "Species", test_options = list(nonparametric=TRUE))
```

## Paired observations

In situations with paired data, the `group` variable usually denotes the timing of the measurement (e.g. "before" and "after" or "time 1", "time 2", etc.). In these scenarios, you need an additional index variable that specifies which observations from the different timepoints should be paired. The `test_options =list(paired=TRUE, indices = <Character name of index variable name or vector of indices>)` option can be used to specify the pairing indices, see the example below. DescrTab2 only works with data in "long format", see e.g. `?reshape` or `?tidyr::pivot_longer` for information on how to transoform your data from wide to long format.

```{r, results='asis'}
descr(dat %>% mutate(animal = fct_recode(animal, Before="Fish", After="Mammal")) %>% select(-"Species"), "animal", test_options = list(paired=TRUE, indices=rep(1:75, each=2)))

descr(dat %>% mutate(animal = fct_recode(animal, Before="Fish", After="Mammal"), idx = rep(1:75, each=2)) %>% select(-"Species"), "animal", test_options = list(paired=TRUE, indices="idx" ))
```

## Significant digits

Every summary statistic in DescrTab2 is formatted by a corresponding formatting function. You can exchange these formatting functions as you please:

```{r, results='asis'}
descr(dat, "Species", format_summary_stats = list(mean=function(x)formatC(x, digits = 4)) )
```


## Omitting summary statistics

Let's say you don't want to calculate quantiles for your numeric variables. You can specify the `summary_stats_cont` option to include all summary statistics but quantiles:

```{r, results='asis'}
descr(dat, "Species", summary_stats_cont = list(N = DescrTab2:::.N, Nmiss = DescrTab2:::.Nmiss, mean =
    DescrTab2:::.mean, sd = DescrTab2:::.sd, median = DescrTab2:::.median, min = DescrTab2:::.min, max =
    DescrTab2:::.max))
```

## Adding summary statistics
Let's say you have a categorical variable, but for some reason it's levels are numerals and you want to calculate the mean. No problem:

```{r, results='asis'}
# Create example dataset
dat2 <- iris
dat2$cat_var <- c(1,2) %>% sample(150, TRUE) %>% factor()
dat2 <- dat2[, c("Species", "cat_var")]

descr(dat2, "Species", summary_stats_cat=list(mean=DescrTab2:::.factormean))

```

## Combining mean and sd
Use the `format_options = list(combine_mean_sd=TRUE)` option:

```{r, results='asis'}
descr(dat, "Species", format_options = c(combine_mean_sd=TRUE))
```


## Omitting p-values
You can declare the `format_options = list(print_p = FALSE)` option to omit p-values:

```{r, results='asis'}
descr(dat, "animal", format_options = list(print_p = FALSE))
```

Similarily for Confidence intervals:

```{r, results='asis'}
descr(dat, "animal", format_options = list(print_CI = FALSE))
```


## Printing without results='asis'
Sometimes, e.g. if you have a loop inside your R-chunk and you want to plot graphics in between descriptive tables, it is necessary not to have the results='asis' option. You can still use DescrTab2 with the following commands:

````markdown
`r ''````{r}
capture.output(print(descr(dat, "Species"))) %>%  knitr::raw_html() # or knitr::raw_tex() for tex
`r ''````
````

```{r}
capture.output(print(descr(dat, "Species"))) %>%  knitr::raw_html() # or knitr::raw_tex() for tex
```

In word documents this is irrelevant, because you never have to specify `results='asis'`


## Controling options on a per-variable level
You can use the `var_options` list to control formatting and test options on a per-variable basis.
Let's say in the dataset `iris`, we want that only the `Sepal.Length` variable has more digits in the mean and a nonparametric test:


```{r, results='asis'}
descr(iris, "Species", var_options = list(Sepal.Length = list(
  format_summary_stats = list(
    mean = function(x)
      formatC(x, digits = 4)
  ),
  test_options = c(nonparametric = TRUE)
)))
```










