---
title: "Usage guidance"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Usage guidance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(knitr)
library(tidyverse)
library(magrittr)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

`DescrTab2` is the replacement of the `DescrTab` package.
It supports a variety of different customization options and can be used
in .Rmd files in conjunction with knitr.

## Preamble settings

### imbi_report
You're all set. Everything is already included.

### pdf_document
Here is what you need to include in the yaml header to use `DescrTab2` inside .Rmd file with pdf_document output:

````markdown
---
title: "DescrTab2 tutorial"
header-includes:
   - \usepackage{needspace}
   - \usepackage{longtable}
   - \usepackage{booktabs}
output: pdf_document
---
````

### html & word_document
No special preamble needed. Make sure you have pandoc version >= 2.0 installed on your system.

## Global print_format option
In order for `DescrTab2` to work properly with your document type of choice, you need to set the `printFormat` options, preferably right at the start of your document. You can do this by typing:

```{r, echo=TRUE, message=FALSE}
options(print_format = "html") # or = "word" or "tex", depending on your document type
```


## Getting started

For instructive purposes, we will use the following dataset:

```{r}
dat <- iris[, c("Species", "Sepal.Length")]
dat %<>% mutate(animal= c("Mammal", "Fish") %>% rep(75) %>% factor())
dat %<>% mutate(food= c("fries", "wedges") %>% sample(150, T) %>% factor())
```


Make sure you include the DescrTab library by typing 

```{r, echo=TRUE}
library(DescrTab2)
```


somewhere in the document before you use it. You are now ready to go! Producing beautiful descriptive tables in html and tex is now as easy as typing:

````markdown
`r ''````{r, results='asis'}
descr(dat)
`r ''````
````

```{r, results='asis', echo=F}
descr(dat)
```


Note the chunk option `results='asis'`. `DescrTab2` produces raw LaTeX or hmtl code. To get pandoc to render this properly, the `results='asis'` option is required. An alternative will be described later.


To produce descriptive tables for a word document, a bit more typing is required:

````markdown
`r ''````{r}
descr(dat) %>% print() %>% knitr::knit_print()
`r ''````
````

When producing word tables in this fashion, you must *not* have the `results='asis'` chunk option set.

Note that `DescrTab2` can also produce console output! In fact, this is the default setting (i.e. if the global `DescrTabFormat` is not specified)


## Accessing table elements

The object returned from the `descr` function is basically just a named list. You may be interested in referencing certain summary statistics from the table in your document. To do this, you can save the list returned by `descr`:

```{r}
my_table <- descr(dat)
```

You can then access the elements of the list using the `$` operator.

```{r}
my_table$variables$Sepal.Length$results$Total$mean
```

Rstudios autocomplete suggestions are very helpful when navigating this list.

The `print` function returns a formatted version of this list, which you can also save and access using the same syntax.

```{r, results='asis'}
my_table <- descr(dat) %>% print(silent=T)
```


## Specifying a group
Use the `group` option to specify the name of a grouping variable in your data:

```{r, results='asis'}
descr(dat, "Species")
```

## Assigning labels
Use the `group_labels` option to assign group labels and the `var_labels` option to assign variable labels:

```{r, results='asis'}
descr(dat, "Species", group_labels=list(setosa="My custom group label"), var_labels = list(Sepal.Length = "My custom variable label"))
```

## Confidence intervals for two group comparisons
For 2-group comparisons, decrtab automatically calculates confidence intervals for differences in effect measures:

```{r, results='asis'}
descr(dat, "animal")
```

## Different tests
There are a lot of different tests available. Check out the test_choice vignette for details: `vignette("test_choice_tree", "DescrTab2")`, or look at https://imbi-heidelberg.github.io/DescrTab2/articles/test_choice_tree_html.html

Here are some different tests in action:

```{r, results='asis'}
descr(dat %>% select(-"Species"), "animal", test_options = list(exact=T, nonparametric=T))
```


```{r, results='asis'}
descr(dat %>% select(c("Species", "Sepal.Length")), "Species", test_options = list(nonparametric=T))
```


```{r, results='asis'}
descr(dat %>% select(-"Species"), "animal", test_options = list(paired=T, indices=rep(1:50, 3)))
```

## Significant digits

Every summary statistic in DescrTab2 is formatted by a corresponding formatting function. You can exchange these formatting functions as you please:

```{r, results='asis'}
descr(dat, "Species", format_summary_stats = list(mean=function(x)formatC(x, digits = 4)) )
```


## Omitting summary statistics

Let's say you don't want to calculate quantiles for your numeric variables. You can specify the `summary_stats_cont` option to include all summary statistics but quantiles:

```{r, results='asis'}
descr(dat, "Species", summary_stats_cont = list(N = DescrTab2:::.N, Nmiss = DescrTab2:::.Nmiss, mean =
    DescrTab2:::.mean, sd = DescrTab2:::.sd, median = DescrTab2:::.median, min = DescrTab2:::.min, max =
    DescrTab2:::.max))
```

## Adding summary statistics
Let's say you have a categorical variable, but for some reason it's levels are numerals and you want to calculate the mean. No problem:

```{r, results='asis'}
# Create example dataset
dat2 <- iris
dat2$cat_var <- c(1,2) %>% sample(150, T) %>% factor()
dat2 <- dat2[, c("Species", "cat_var")]

descr(dat2, "Species", summary_stats_cat=list(mean=DescrTab2:::.factormean))

```

## Combining mean and sd
Use the `format_options = list(combine_mean_sd=T)` option:

```{r, results='asis'}
descr(dat, "Species", format_options = c(combine_mean_sd=T))
```


## Omitting p values

You can declare the `format_options = list(print_p = F)` option to omit p-values:

```{r, results='asis'}
descr(dat, "animal", format_options = list(print_p = F))
```

Similarily for Confidence intervals:

```{r, results='asis'}
descr(dat, "animal", format_options = list(print_CI = F))
```


## Printing without results='asis'

Sometimes, e.g. if you have a loop inside your R-chunk and you want to plot graphics in between descriptive tables, it is necessary not to have the results='asis' option. You can still use DescrTab2 with the following commands:

````markdown
`r ''````{r}
capture.output(print(descr(dat, "Species"))) %>%  knitr::raw_html() # or knitr::raw_tex() for tex
`r ''````
````

```{r}
capture.output(print(descr(dat, "Species"))) %>%  knitr::raw_html() # or knitr::raw_tex() for tex
```

In word documents this is irrelevant, because you never have to specify `results='asis'`


## Controling options on a per-variable level
You can use the `var_options` list to control formatting and test options on a per-variable basis.
Let's say in the dataset `iris`, we want that only the `Sepal.Length` variable has more digits in the mean and a nonparametric test:


```{r, results='asis'}
descr(iris, "Species", var_options = list(Sepal.Length = list(
  format_summary_stats = list(
    mean = function(x)
      formatC(x, digits = 4)
  ),
  test_options = c(nonparametric = T)
)))
```










