---
title: "DescrTab2 usage guidance"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Usage guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(knitr)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

`DescrTab2` is the replacement of the `DescrTab` package.
It supports a variety of different customization options and can be used
in .Rmd files in conjunction with knitr.

## Preamble settings

### imbi_report
You're all set. Everything is already included.

### pdf_document
Here is what you need to include in the yaml header to use `DescrTab2` inside .Rmd file with pdf_document output:

````markdown
---
title: "DescrTab2 tutorial"
header-includes:
   - \usepackage{needspace}
   - \usepackage{longtable}
   - \usepackage{booktabs}
output: pdf_document
---
````

### html & word_document
No special preamble needed. Make sure you have pandoc version >= 2.0 installed on your system.

## Global printFormat option
In order for `DescrTab2` to work properly with your document type of choice, you need to set the `printFormat` options, preferably right at the start of your document. You can do this by typing:

```{r, echo=TRUE, message=FALSE}
options(DescrTabFormat = "html") # or = "word" or "tex", depending on your document type

```


## Getting started

Make sure you include the DescrTab library by typing 


```{r, echo=TRUE}
# library(DescrTab2)
```


somewhere in the document before you use it. You are now ready to go! Producing beautiful descriptive tables in html and tex is now as easy as typing:

````markdown
`r ''````{r, results='asis'}
descr(iris)
`r ''````
````

Note the chunk option `results='asis'`. `DescrTab2` produces raw LaTeX or hmtl code. To get pandoc to render this properly, the `results='asis'` option is required. An alternative will be described later.


To produce descriptive tables for a word document, a bit more typing is required:

````markdown
`r ''````{r}
descr(iris) %>% print() %>% .$ft %>% knitr::knit_print()
`r ''````
````

When producing word tables in this fashion, you must *not* have the `results='asis'` chunk option set.

Note that `DescrTab2` can also produce console output! In fact, this is the default setting (i.e. if the global `DescrTabFormat` is not specified)


## Accessing table elements

The object returned from the `descr` function is basically just a named list. You may be interested in referencing certain summary statistics from the table in your document. To do this, you can save the list returned by `descr`:

```{r}
# my_table <- descr(iris)
```

You can then access the elements of the list using the `$` operator.

```{r}
# my_table$variables$Sepal.Length$Total$mean
```

Rstudios autocomplete suggestions are very helpful when navigating this list.

The `print` function returns a formatted version of this list, which you can also save and access using the same syntax.

```{r}
# my_table <- descr(iris) %>% print()
```


## Specifying a group
Use the `group` option to specify the name of a grouping variable in your data:

```{r, results='asis'}
# descr(iris, "Species")
```

## Assigning labels
Use the `group_labels` option to assign group labels and the `var_options` option to assign variable labels:

```{r, results='asis'}
# descr(iris, "Species", group_labels=list(setosa="My custom group label"), var_options = list(Sepal.Length = list(label="My custom variable label")))
```



